---
title: "Group 14 Project"
author: "Edison Cheah, Dev Nair, Eldrick Chen"
date: "04/07/2026"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r better_notation}
 options(scipen = 999)
```

```{r message=FALSE}
library(dplyr)

# Load in March Madness datasets
original_dataset <- read.csv("cbb.csv")
dataset1 <- read.csv("march-madnesss-historical-data.csv")
dataset2 <- read.csv("Int _ KenPom _ Height.csv")

# DATA CLEANING 1: Purge all non-tournament teams from dataset
non_purged_data <- original_dataset[, c("TEAM", "POSTSEASON")]

original_dataset <- original_dataset %>%
  filter(!is.na(POSTSEASON) & POSTSEASON != "NA" & POSTSEASON != "N/A")

purged_data <- original_dataset[, c("TEAM", "POSTSEASON")]

# DATA CLEANING 2: Standardize team names
unstandardized_large_teams <- sort(unique(dataset1$Team))

original_dataset$TEAM <- gsub("[^A-Za-z ]", "", original_dataset$TEAM)
dataset1$Team <- gsub("[^A-Za-z ]", "", dataset1$Team)

dataset1$Team <- gsub("State", "St.", dataset1$Team)

dataset1$Team[dataset1$Team == "Brigham Young"] <- "BYU"
dataset1$Team[dataset1$Team == "FDU"] <- "Fairleigh Dickinson"
dataset1$Team[dataset1$Team == "GardnerWebb"] <- "Gardner Webb"
dataset1$Team[dataset1$Team == "Louisiana"] <- "Louisiana Lafayette"
dataset1$Team[dataset1$Team == "Louisiana St."] <- "LSU"
dataset1$Team[dataset1$Team == "Loyola IL"] <- "Loyola Chicago"
dataset1$Team[dataset1$Team == "North Carolina AT"] <- "North Carolina A&T"
dataset1$Team[dataset1$Team == "Prairie View"] <- "Prairie View A&M"
dataset1$Team[dataset1$Team == "Saint Josephs"] <- "Saint Joseph's"
dataset1$Team[dataset1$Team == "Saint Marys"] <- "Saint Mary's"
dataset1$Team[dataset1$Team == "Saint Peters"] <- "Saint Peter's"
dataset1$Team[dataset1$Team == "St Johns NY"] <- "St John's"
dataset1$Team[dataset1$Team == "Stephen F Austin"] <- "Stephen F. Austin"
dataset1$Team[dataset1$Team == "Texas AM"] <- "Texas A&M"
dataset1$Team[dataset1$Team == "Texas AMCorpus Christi"] <- "Texas A&M Corpus Chris"

dataset2$Team[dataset2$Team == "St. Bonaventure"] <- "St Bonaventure"
dataset2$Team[dataset2$Team == "Mount St. Mary's"] <- "Mount St Marys"

# Create list of unique teams
original_teams_unique <- sort(unique(original_dataset$TEAM))
ds1_teams_unique <- sort(unique(dataset1$Team))
ds2_teams_unique <- sort(unique(dataset2$Team))
merged_teams_unique <- intersect(original_teams_unique, ds1_teams_unique)
merged_teams_unique <- intersect(merged_teams_unique, ds2_teams_unique)

# Filter data to only include common teams/years
dataset1 <- dataset1 %>%
  filter(Team %in% merged_teams_unique & !(Year %in% c(2012, 2024)))
original_dataset <- original_dataset %>%
  filter(TEAM %in% merged_teams_unique & !(YEAR %in% c(2012, 2024)))

# DATA CLEANING 3: Merge datasets based on team and year
merged_dataset <- merge(original_dataset,
                        dataset1[, c("Team", 
                                     "Year", 
                                     "FG.",
                                     "OppO",
                                     "OppO",
                                     "FT.",
                                     "Made.Round.of.16",
                                     "Conf..W",
                                     "Conf..L",
                                     "Home.W",
                                     "Home.L",
                                     "Away.W",
                                     "Away.L",
                                     "SOS")], 
                        by.x = c("TEAM", "YEAR"),
                        by.y = c("Team", "Year")
)

merged_dataset <- merge(merged_dataset,
                        dataset2[, c("TeamName", 
                                     "Season", 
                                     "AvgHeight")], 
                        by.x = c("TEAM", "YEAR"),
                        by.y = c("TeamName", "Season")
)

# Write out full conference names
merged_dataset$CONF <- factor(merged_dataset$CONF,
                              levels=unique(sort(merged_dataset$CONF)),
                              labels=c("Atlantic 10",
                                       "Atlantic Coast Conference",
                                       "America East",
                                       "American",
                                       "ASUN",
                                       "Big Ten",
                                       "Big 12",
                                       "Big East",
                                       "Big Sky",
                                       "Big South",
                                       "Big West",
                                       "Colonial Athletic Association",
                                       "Conference USA",
                                       "Horizon League",
                                       "Ivy League",
                                       "Metro Atlantic Athletic Conference",
                                       "Mid-American Conference",
                                       "Mid-Eastern Athletic Conference",
                                       "Missouri Valley Conference",
                                       "Mountain West",
                                       "Northeast Conference",
                                       "Ohio Valley Conference",
                                       "Pac-12",
                                       "Patriot League",
                                       "Sun Belt",
                                       "Southern Conference",
                                       "South Eastern Conference",
                                       "Southland Conference",
                                       "Summit League",
                                       "Southwestern Athletic Conference",
                                       "Western Athletic Conference",
                                       "West Coast Conference"
                                       ))

# DATA CLEANING 4: Create new variable OppNetRtg (net rating)
merged_dataset$OppNetRtg <- merged_dataset$OppO - merged_dataset$OppO

merged_dataset$POSTSEASON <- factor(merged_dataset$POSTSEASON, 
                                    labels=c("R68", "R64", "R32", "S16", "E8", "F4", "2ND", "Champions"), 
                                    ordered=TRUE)

# Calculate win %s
merged_dataset$HomeWinPct <- round(merged_dataset$Home.W / (merged_dataset$Home.W + merged_dataset$Home.L), 3)
merged_dataset$AwayWinPct <- round(merged_dataset$Away.W / (merged_dataset$Away.W + merged_dataset$Away.L), 3)
merged_dataset$ConfWinPct <- round(merged_dataset$Conf..W / (merged_dataset$Conf..W + merged_dataset$Conf..L), 3)
merged_dataset$OvrWinPct <- round(merged_dataset$W / merged_dataset$G, 3)

#Subset based on Sweet 16 qualification
teams_made_S16 <- merged_dataset[which(merged_dataset$Made.Round.of.16 == 1), ]
teams_no_S16 <- merged_dataset[which(merged_dataset$Made.Round.of.16 == 0), ]

# Two sided t-tests
t.test(teams_made_S16$HomeWinPct, teams_no_S16$HomeWinPct, "two.sided")
t.test(teams_made_S16$AwayWinPct, teams_no_S16$AwayWinPct, "two.sided")
t.test(teams_made_S16$ConfWinPct, teams_no_S16$ConfWinPct, "two.sided")
t.test(teams_made_S16$OvrWinPct, teams_no_S16$OvrWinPct, "two.sided")

# Box plots of success vs records
library(ggplot2)
theme_update(text=element_text(size=14))

ggplot(merged_dataset, aes(HomeWinPct, factor(Made.Round.of.16, labels=c("No", "Yes")))) + 
  geom_boxplot() +
  labs(title="Home Win % Among Sweet 16 vs. Non-Sweet 16 Teams",
       x="Win % (Home Games)", 
       y="Whether Team Reached Sweet 16")

ggplot(merged_dataset, aes(AwayWinPct, factor(Made.Round.of.16, labels=c("No", "Yes")))) + 
  geom_boxplot() +
  labs(title="Away Win % Among Sweet 16 vs. Non-Sweet 16 Teams",
       x="Win % (Away Games)",
       y="Whether Team Reached Sweet 16")

ggplot(merged_dataset, aes(ConfWinPct, factor(Made.Round.of.16, labels=c("No", "Yes")))) + 
  geom_boxplot() +
  labs(title="Conference Win % Among Sweet 16 vs. Non-Sweet 16 Teams",
       x="Win % (Conference Games)", 
       y="Whether Team Reached Sweet 16")

ggplot(merged_dataset, aes(OvrWinPct, factor(Made.Round.of.16, labels=c("No", "Yes")))) + 
  geom_boxplot() +
  labs(title="Overall Win % Among Sweet 16 vs. Non-Sweet 16 Teams",
       x="Win % (Overall Games)", 
       y="Whether Team Reached Sweet 16")

# Analyzing 3PT%, 2PT%, FG%, FT%
t.test(teams_made_S16$X3P_O, teams_no_S16$X3P_O, "two.sided")
t.test(teams_made_S16$X2P_O, teams_no_S16$X2P_O, "two.sided")
t.test(teams_made_S16$FG.*100, teams_no_S16$FG.*100, "two.sided")
t.test(teams_made_S16$FT., teams_no_S16$FT., "two.sided")

# Analyzing TOR, TORD
t.test(teams_made_S16$TOR, teams_no_S16$TOR, "two.sided")

ggplot(merged_dataset, aes(X3P_O, factor(Made.Round.of.16, labels=c("No", "Yes")))) + 
  geom_boxplot() +
  labs(title="3P% Among Sweet 16 vs. Non-Sweet 16 Teams",
       x="Three Point %", 
       y="Whether Team Reached Sweet 16")

ggplot(merged_dataset, aes(X2P_O, factor(Made.Round.of.16, labels=c("No", "Yes")))) + 
  geom_boxplot() +
  labs(title="2P% Among Sweet 16 vs. Non-Sweet 16 Teams",
       x="Two Point %", 
       y="Whether Team Reached Sweet 16")

ggplot(merged_dataset, aes(FG., factor(Made.Round.of.16, labels=c("No", "Yes")))) + 
  geom_boxplot() +
  labs(x="Field Goal %", y="Whether Team Reached Sweet 16")

ggplot(merged_dataset, aes(FT.*100, factor(Made.Round.of.16, labels=c("No", "Yes")))) + 
  geom_boxplot() +
  labs(title="FT% Among Sweet 16 vs. Non-Sweet 16 Teams",
       x="Free Throw %", 
       y="Whether Team Reached Sweet 16")

# Analyzing TOR vs made S16
ggplot(merged_dataset, aes(TOR, factor(Made.Round.of.16, labels=c("No", "Yes")))) + 
  geom_boxplot() +
  labs(title="TO% Among Sweet 16 vs. Non-Sweet 16 Teams",
       x="Turnover %", 
       y="Whether Team Reached Sweet 16")

# Analyzing defensive metrics
t.test(teams_made_S16$ADJDE, teams_no_S16$ADJDE, "two.sided")
t.test(teams_made_S16$X3P_D, teams_no_S16$X3P_D, "two.sided")
t.test(teams_made_S16$X2P_D, teams_no_S16$X2P_D, "two.sided")
t.test(teams_made_S16$TORD, teams_no_S16$TORD, "two.sided")
t.test(teams_made_S16$OppO, teams_no_S16$OppO, "two.sided")

ggplot(merged_dataset, aes(ADJDE, factor(Made.Round.of.16, labels=c("No", "Yes")))) + 
  geom_boxplot() +
  labs(title="AdjDE Among Sweet 16 vs. Non-Sweet 16 Teams",
       x="Adjusted Defensive Efficiency", 
       y="Whether Team Reached Sweet 16")

ggplot(merged_dataset, aes(X3P_D, factor(Made.Round.of.16, labels=c("No", "Yes")))) + 
  geom_boxplot() +
  labs(title="Opponent 3P% Among Sweet 16 vs. Non-Sweet 16 Teams",
       x="Opponent Three Point %", 
       y="Whether Team Reached Sweet 16")

ggplot(merged_dataset, aes(X2P_D, factor(Made.Round.of.16, labels=c("No", "Yes")))) + 
  geom_boxplot() +
  labs(title="Opponent 2P% Among Sweet 16 vs. Non-Sweet 16 Teams",
       x="Opponent Two Point %",
       y="Whether Team Reached Sweet 16")

ggplot(merged_dataset, aes(TORD, factor(Made.Round.of.16, labels=c("No", "Yes")))) + 
  geom_boxplot() +
  labs(title="Opponent TO% Among Sweet 16 vs. Non-Sweet 16 Teams",
       x="Opponent Turnover %",
       y="Whether Team Reached Sweet 16")

ggplot(merged_dataset, aes(OppO, factor(Made.Round.of.16, labels=c("No", "Yes")))) + 
  geom_boxplot() +
  labs(x="Opponent Defensive Rating", y="Whether Team Reached Sweet 16")

# Analyzing height
AdjDE_vs_height <- lm(merged_dataset$ADJDE ~ merged_dataset$AvgHeight)
summary(OppO_vs_height)

ggplot(merged_dataset, aes(x = AvgHeight, y = ADJDE, color = factor(Made.Round.of.16))) +
  geom_point(alpha=0.2) +
  geom_smooth(method = "lm") +
  scale_color_manual(values = c("0" = "red", "1" = "blue"),
                     labels = c("0" = "No", "1" = "Yes")) +
  labs(
    title = "AdjDE vs. Average Height by Sweet 16 Qualification",
    x = "Average Height",
    y = "Adjusted Defensive Efficiency",
    color = "Made Sweet 16"
  )

EFG_D_vs_height <- lm(merged_dataset$EFG_D ~ merged_dataset$AvgHeight)
summary(EFG_D_vs_height)

ggplot(merged_dataset, aes(x = AvgHeight, y = EFG_D, color = factor(Made.Round.of.16))) +
  geom_point(alpha=0.2) +
  geom_smooth(method = "lm") +
  scale_color_manual(values = c("0" = "red", "1" = "blue"),
                     labels = c("0" = "No", "1" = "Yes")) +
  labs(
    title = "Opponent EFG% vs. Average Height by Sweet 16 Qualification",
    x = "Average Height",
    y = "Opponent Effective FG%",
    color = "Made Sweet 16"
  )

t.test(teams_made_S16$AvgHeight, teams_no_S16$AvgHeight, "two.sided")

ggplot(merged_dataset, aes(AvgHeight, factor(Made.Round.of.16, labels=c("No", "Yes")))) + 
  geom_boxplot() +
  labs(x="Average Height (in)", y="Whether Team Reached Sweet 16")

t.test(teams_made_S16$SOS, teams_no_S16$SOS, "two.sided")

ggplot(merged_dataset, aes(SOS, factor(Made.Round.of.16, labels=c("No", "Yes")))) + 
  geom_boxplot() +
  labs(x="Strength of Schedule", y="Whether Team Reached Sweet 16")

contingency_table <- table(merged_dataset$CONF, merged_dataset$Made.Round.of.16)
chisq.test(contingency_table)
```

# Data Cleaning Identification and Justification

In order to properly use our March Madness datasets, they need to be standardized and made relevant. This section outlines the three data cleaning steps we took to perform analysis on our data.

1. The original dataset provided was purged of non-tournament teams. Since most of the March Madness datasets we found only included teams who qualified for the postseason, a more robust analysis can be performed by only looking at in-tournament teams. Also, it allows us to focus on our pending question of what stats define postseason success for a team. Table 1 shows a preview of the data before and after this step.

`r knitr::kable(list(non_purged_data[50:60, ], purged_data[50:60, ]), caption = "Comparison of before (left) vs. after (right) purging non-tournament teams")`

2. Team names were standardized between datasets, following the convention of the original dataset. By doing this, fields across datasets can be joined based on common Team and Year fields upon merging. Table 2 is a preview of the data before and after this step.

`r knitr::kable(list(unstandardized_large_teams[c(15, 21, 22, 44, 98)], large_teams_unique[c(15, 21, 22, 44, 98)]), caption = "Comparison of before (left) vs. after (right) standardizing names")`

3. The datasets were merged based on commonalities of Team and Year, with the final dataset being the original dataset with the supplementary dataset's columns: OppO, OppO, Made.Round.of.16. With these columns, we can calculate another performance metric (net rating) and better assess postseason success. Additional columns may be merged in the future to expand upon our data analysis. Table 3 shows a comparison of column names from the original dataset and the merged dataset.

`r knitr::kable(list(colnames(original_dataset), colnames(merged_dataset)), caption = "Comparison of column names before (left) vs. after (right) merging datasets")`
